---
title: Heatmaps of genes of interest 
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  BiocStyle::html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(magrittr))
suppressMessages(library(kableExtra))
suppressMessages(library(ggplot2))
suppressMessages(library(data.table))
suppressMessages(library(ggrepel))
suppressMessages(library(limma))
suppressMessages(library(edgeR))
suppressMessages(library(dplyr))
suppressMessages(library(genefilter))
suppressMessages(library(RColorBrewer))
suppressMessages(library(gridExtra))
suppressMessages(library(VennDiagram))
suppressMessages(library(ggthemes))
suppressMessages(library(pheatmap))
futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")
```




```{r readInData, message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE,  context="data"}
counts <- read.table(file = "gene.counts.txt", header = TRUE, check.names=FALSE, row.names=1)
decoderFile <- "decoder.txt"
decoder.data <- fread(decoderFile) %>% as.data.frame()
decoder.data$group <- factor(make.names(decoder.data$group))
decoder.data$Treatment <- factor(decoder.data$toxin, levels=c("PBS", "ETX", "PTX"))
decoder.data$tissue <- factor(decoder.data$tissue)
decoder.data <- decoder.data[decoder.data$sample.ID %in% colnames(counts),]
counts <- counts[,c(decoder.data$sample.ID)]
#table(colnames(counts) == decoder.data$sample.ID )
```


## Samples

The following samples were part of this analysis:

```{r samples, message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE,  context="data"}
 kable(decoder.data[,c(-1)], row.names=FALSE,  padding = 0, longtable=TRUE) %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width=F)
```



# Differential expression analysis

Based on the samples, the following pairwise comparisons were made via limma-voom:

- spinal cord PTX vs. spinal cord (no Treatment)
- spinal cord ETX vs. spinal cord (no Treatment)


```{r runLimma,  eval=T, message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE}
dge <- DGEList(counts=counts)
design = model.matrix(~ 0 + group, decoder.data)
colnames(design) <- gsub("group", "", colnames(design))
keep <- filterByExpr(dge, design)
dge <- dge[keep,,keep.lib.sizes=FALSE]
dge <- calcNormFactors(dge)
normCounts = cpm(dge, log=T)
v <- voom(dge, design, plot=F)
fit <- lmFit(v, design)
cont.matrix <- makeContrasts(PTX.SC-CT.SC,
                             ETX.SC-CT.SC,
                             levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

PTX.SC_vs_CT.SC <- topTable(fit2,coef=1,number=Inf)
ETX.SC_vs_CT.SC <- topTable(fit2,coef=2,number=Inf)
```


The following genes were detected as differentially expresed (FDR < 0.10):

```{r de_res, eval=T,  warning=FALSE, cache=TRUE, echo=FALSE}
de.all = list(PTX.SC_vs_CT.SC = PTX.SC_vs_CT.SC,
              ETX.SC_vs_CT.SC = ETX.SC_vs_CT.SC)
alpha = 0.10

de.all.sig <- lapply(de.all, subset, adj.P.Val < alpha)

res =  as.data.frame(sapply(de.all.sig, nrow))
colnames(res) <- paste0("padj<",alpha)


all.comps = stringr::str_split(row.names(res), "_vs_")

rel.contrs = 1:length(all.comps)
 
kable(data.frame(res[rel.contrs,, drop=F], contr=rel.contrs), row.names=T,  padding = 0, longtable=TRUE) %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width=F)
```


#  Heatmaps of genes of interest (Fig. 7E)

heat map displaying genes of interest reveals induction of protease, signal transduction, cytokine, and transcription factor genes relevant to overcoming CNS immune privilege 

```{r hm1, message=FALSE, warning=FALSE, eval=TRUE, cache=TRUE, echo=FALSE,  context="data", fig.align='center', fig.width=6, fig.height=6}

goi = c("Ibtk","Vegfa","Pla2g7","Mmp2","Mmp25","Mmp28","Mmp17","Adamtsl1","Nup205","Cxcl16","Cxcl12","Tnfrsf22","Pik3ca","Adamts8","Adamts2","Adamts6","Adamts20","Ripk2","Casp7","Casp9","Casp2","Casp8","Timp2","Tlr3","Tlr4","Tlr2","IL17rb","IL17ra","IL27ra","IL3ra","Socs4","IL6","Irak4","Cd14","Traf1","Tirap","Notch2","Akt2","Pik3ca","Cdh2")
goi = stringr::str_to_sentence(goi)
#table(goi %in% row.names(counts))

is_de = unique(unlist(lapply(de.all.sig, row.names)))

goi = goi[goi %in% is_de]
decod.sub = decoder.data
annot_col =data.frame(Treatment=decod.sub$Treatment, row.names=decod.sub$sample.ID)
samps = decod.sub$sample.ID
log2.count.matrix = normCounts[goi,samps] %>% as.matrix()

annoCol<-list(Treatment=c(PBS=rgb(142,78,137,maxColorValue = 255), ETX=rgb(201,66,53,maxColorValue = 255), PTX=rgb(84,149,78,maxColorValue = 255)))
cols <- rev(colorRampPalette(brewer.pal(10, "RdBu"))(50))

pheatmap(log2.count.matrix, scale="row", show_rownames=T, fontsize = 10, cellwidth =15, cellheight=9, main = "",  annotation_col = annot_col,color=cols, silent = F ,  cluster_cols = T, show_colnames = F, treeheight_col=0, treeheight_row=20,  annotation_colors = annoCol)
```





#  Heatmaps of genes of interest (Fig. 7f)

we show a heat map of genes previously identified by Munji et al as components of a core set of genes involved in BBB dysfunction 

(asterik = DE in one of the contrasts)

```{r hm_2, message=FALSE, warning=FALSE, eval=TRUE, cache=TRUE, echo=FALSE,  context="data", fig.align='center', fig.width=6, fig.height=6}

goi = c("Plekho1","Mycl","Sele","Upp1","Adamts8","Lamb1","Lrg1","Plekho2","Scgb3a1","Selp","Col3a1","Serpine1","Serping1","Serpina3g","Cenpa")
goi = stringr::str_to_sentence(goi)

is_de = unique(unlist(lapply(de.all.sig, row.names)))

goi = goi[goi %in% is_de]
decod.sub = decoder.data
annot_col =data.frame(Treatment=decod.sub$Treatment, row.names=decod.sub$sample.ID)
samps = decod.sub$sample.ID
log2.count.matrix = normCounts[goi,samps] %>% as.matrix()

pheatmap(log2.count.matrix, scale="row", show_rownames=T, fontsize = 10, cellwidth =15, cellheight = 15, main = "",  annotation_col = annot_col,color=cols, silent = F , cluster_cols = T, show_colnames = F, treeheight_col=0, treeheight_row=20, annotation_colors = annoCol)
```






# Session Info
```{r session, message=FALSE, warning=FALSE, cache=TRUE,comment="#",echo=FALSE, fig.width=10, fig.height=5.5, context="data"}
sessionInfo()
```
