---
title: Venn + Scatterplot 
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  BiocStyle::html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(magrittr))
suppressMessages(library(kableExtra))
suppressMessages(library(ggplot2))
suppressMessages(library(data.table))
suppressMessages(library(ggrepel))
suppressMessages(library(limma))
suppressMessages(library(edgeR))
suppressMessages(library(dplyr))
suppressMessages(library(genefilter))
suppressMessages(library(RColorBrewer))
suppressMessages(library(gridExtra))
suppressMessages(library(VennDiagram))
suppressMessages(library(ggthemes))
futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")
```




```{r readInData, message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE,  context="data"}
counts <- read.table(file = "gene.counts.txt", header = TRUE, check.names=FALSE, row.names=1)
decoderFile <- "decoder.txt"
decoder.data <- fread(decoderFile) %>% as.data.frame()
decoder.data$group <- factor(make.names(decoder.data$group))
decoder.data$toxin <- factor(decoder.data$toxin, levels=c("PBS", "ETX", "PTX"))
decoder.data$tissue <- factor(decoder.data$tissue)
decoder.data <- decoder.data[decoder.data$sample.ID %in% colnames(counts),]
counts <- counts[,c(decoder.data$sample.ID)]
#table(colnames(counts) == decoder.data$sample.ID )
```


## Samples

The following samples were part of this analysis:

```{r samples, message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE,  context="data"}
 kable(decoder.data[,c(-1)], row.names=FALSE,  padding = 0, longtable=TRUE) %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width=F)
```



# Differential expression analysis

Based on the samples, the following pairwise comparisons were made via limma-voom:

- spinal cord PTX vs. spinal cord (no toxin)
- spinal cord ETX vs. spinal cord (no toxin)


```{r runLimma,  eval=T, message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE}
dge <- DGEList(counts=counts)
design = model.matrix(~ 0 + group, decoder.data)
colnames(design) <- gsub("group", "", colnames(design))
keep <- filterByExpr(dge, design)
dge <- dge[keep,,keep.lib.sizes=FALSE]
dge <- calcNormFactors(dge)
normCounts = cpm(dge, log=T)
v <- voom(dge, design, plot=F)
fit <- lmFit(v, design)
cont.matrix <- makeContrasts(PTX.SC-CT.SC,
                             ETX.SC-CT.SC,
                             levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

PTX.SC_vs_CT.SC <- topTable(fit2,coef=1,number=Inf)
ETX.SC_vs_CT.SC <- topTable(fit2,coef=2,number=Inf)
```


The following genes were detected as differentially expresed (FDR < 0.10):

```{r de_res, eval=T,  warning=FALSE, cache=TRUE, echo=FALSE}
de.all = list(PTX.SC_vs_CT.SC = PTX.SC_vs_CT.SC,
              ETX.SC_vs_CT.SC = ETX.SC_vs_CT.SC)
alpha = 0.10

de.all.sig <- lapply(de.all, subset, adj.P.Val < alpha)

res =  as.data.frame(sapply(de.all.sig, nrow))
colnames(res) <- paste0("padj<",alpha)


all.comps = stringr::str_split(row.names(res), "_vs_")

rel.contrs = 1:length(all.comps)
 
kable(data.frame(res[rel.contrs,, drop=F], contr=rel.contrs), row.names=T,  padding = 0, longtable=TRUE) %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width=F)
```


# Venn diagrams comparing PTX.SC_vs_CT.SC	 and ETX.SC_vs_CT.SC


```{r venn, message=FALSE, warning=FALSE, eval=TRUE, cache=TRUE, echo=FALSE,  context="data", fig.align='center', fig.width=5, fig.height=5}
library(venneuler)
library(eulerr)

ll = list("ETX vs. PBS" = row.names(subset(ETX.SC_vs_CT.SC, adj.P.Val < alpha)), "PTX vs. PBS" = row.names(subset(PTX.SC_vs_CT.SC, adj.P.Val < alpha)))

p = plot(euler(ll),
     fills =c(rgb(201,66,53,maxColorValue = 255), rgb(84,149,78,maxColorValue = 255)),
     alpha=0.7,
     edges = T,
     quantities = list(fontsize = 18, just="center"), 
     labels= list(fontsize = 18,  just="center"), adjust_labels=T)

grid.arrange(p,ncol=1,padding=unit(0.9,"line"),top="",bottom="",right="",left="")

```


## Scatter plot comparing fold-changes of genes that are in the intersection


```{r scatterplot, message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE, fig.width=6, fig.height=6, context="data",context="data"}
PTX.SC_vs_CT.SC_sig = subset(PTX.SC_vs_CT.SC, adj.P.Val < alpha)
colnames(PTX.SC_vs_CT.SC_sig) <- paste0("PTX.SC_vs_CT.SC_",colnames(PTX.SC_vs_CT.SC_sig))

ETX.SC_vs_CT.SC_sig = subset(ETX.SC_vs_CT.SC, adj.P.Val < alpha)
colnames(ETX.SC_vs_CT.SC_sig) <- paste0("ETX.SC_vs_CT.SC_",colnames(ETX.SC_vs_CT.SC_sig))

merged_dfs_all = merge(PTX.SC_vs_CT.SC_sig, ETX.SC_vs_CT.SC_sig, by="row.names", all="TRUE")

merged_dfs = merge(PTX.SC_vs_CT.SC_sig, ETX.SC_vs_CT.SC_sig, by="row.names")
merged_dfs$meanLFC = rowMeans(merged_dfs[,c("PTX.SC_vs_CT.SC_logFC","ETX.SC_vs_CT.SC_logFC")])
merged_dfs = merged_dfs[order(merged_dfs$meanLFC),]
merged_dfs$gene = merged_dfs$Row.names
merged_dfs$direction = ifelse(merged_dfs$meanLFC > 0, "up", "down")

library(ggpubr)
library(ggsci)


ggplot(merged_dfs, aes(x=ETX.SC_vs_CT.SC_logFC, y=PTX.SC_vs_CT.SC_logFC, label=gene)) + geom_point(size=3, alpha=0.6, pch=21, color="black", aes(fill=direction)) + theme_bw(base_size = 16) + theme(legend.position =  "bottom") + scale_fill_aaas(name=bquote(log[2]~fold~change~direction)) + ggtitle("Scatter plot of fold-changes", subtitle="comparing changes induced in ETX/PTX relative to PBS") + xlab(bquote(log[2]~FC~(ETX - PBS))) +  ylab(bquote(log[2]~FC~(PTX - PBS))) +  stat_cor( aes(label = paste(..rr.label.., sep = "~`,`~")),label.x = -7) + geom_smooth(method=lm, se=T) + geom_vline(xintercept = 0, linetype="dotted", size=1)  + geom_hline(yintercept = 0, linetype="dotted", size=1)  +  theme(plot.subtitle=element_text(size=14, color="black"))
```


# Session Info
```{r session, message=FALSE, warning=FALSE, cache=TRUE,comment="#",echo=FALSE, fig.width=10, fig.height=5.5, context="data"}
sessionInfo()
```
